---
title: "Distribution of tuberculosis-like lesions in wild boar"
format:
  html:
    toc: true
    html-math-method: katex
    embed-resources: true
    monobackgroundcolor: "white"
knitr:
  opts_chunk: 
    echo: true
    fig.align: center
highlight-style: github-dark
---

In this activity, you will practice fitting a generalized linear model. The data are from Vicente *et al* (2006), who analysed the distribution of tuberculosis-like lesions in wild boar, *Sus scrofa*. Part of this study was looking at the potential importance of wild board in maintaining tuberculosis in south central Spain.

Before we being, let's load some packages

```{r}
#| message: false
#| label: setup
library("janitor")
library("dplyr")
library("tibble")
library("tidyr")
library("readr")
library("ggplot2")
library("gglm")
library("performance")
library("marginaleffects")
```

Now we can load the data

```{r}
wb <- read_table("data/boar.txt", col_types = "dddd") |>
  janitor::clean_names()
```

There are four variables in the data set:

* `tb` and indicator of the presence (`1`) or absence (`0`) of tuberculoisis-like lesions,
* `sex` coded as `1` or `2`,
* `age_class`, and
* `length_ct`, which is the length of the animal's head and body.

We should recode some of these variables if we intended to use `sex` and `age_class` in models as these are categorical variables and are best coded as factors in R. For plotting it can also be useful to do this:

```{r}
wb <- wb |>
  mutate(
    f_tb = factor(tb, levels = c(0,1))
  )
```

Start by producing boxplots of `length_ct` against `tb` presence and absence

```{r}
wb |>
  ggplot(
    aes(
      y = length_ct, x = f_tb, fill = f_tb
    )
  ) +
  geom_boxplot() +
  labs(
    x = "TB status",
    fill = "TB Status",
    y = "Length CT"
  )
```

There is a tendency, in the data for animals with tubercolosis-like lesions to be larger. Of course, in any real data analysis, we should investigate why there are missing values in the `tb` variable and account for potential missing data problems.

::: {.callout-note}
# Question

However, for now, let's proceed by fitting a simple binomial GLM to these data. The model we want to fit is

\begin{align*}
\mathtt{tb}_i \sim& \mathcal{B}(\mu_i, 1) \\
\mathbb{E}(\mathtt{tb}_i) =& \mu_i \\
\text{logit}(\mu_i) =& \beta_0 + \beta_1 \mathtt{length\_ct}
\end{align*}

The response variable is `tb` (or `f_tb` can be used), and the predictor variable is `length_ct`. Fit this model in R.
:::

::: {.callout-tip collapse="true"}
# Answer

```{r fit-tb-glm}
wb_glm <- glm(
  tb ~ length_ct,
  data = wb,
  family = binomial(link = "logit")
)
```

:::

::: {.callout-note}
# Question

Produce the model summary. Is the effect of `length_ct` on the prevalence of tuberculosis-like lesions statistically interesting?

:::

::: {.callout-tip collapse="true"}
# Answer

```{r fsummary-tb-glm}
summary(wb_glm)
```

Yes, the effect is statistically interesting; the *p* values is small.

:::

Having fitted a model, it is often useful to plot estimates from the model at representative values of the covariates. The *marginaleffects* package makes this easy and forces us to condition on specific values of the variables in more complex situations.

Here, we just prepare a plot of the fitted model `wb_glm` using `plot_predictions()`:

```{r}
wb_glm |> plot_predictions(condition = "length_ct")
```

We can add the original data to this plot, as a point layer

```{r}
wb_glm |> plot_predictions(condition = "length_ct") +
  geom_point(
    data = wb,
    aes(x = length_ct, y = tb),
    position = position_jitter(height = 0.025),
    alpha = 0.5
  )
```

::: {.callout-note}
# Question

Look at the model diagnostics. Are there any potential problems with the model?

:::

::: {.callout-tip collapse="true"}
# Answer

```{r gglm-diagnostics-tb-glm}
gglm(wb_glm)
```

Err...

That was a trick question; model diagnostics are notoriously difficult for binary responses. Better diagnostics can be achieved using the *performance* package and it's `check_model()` function:

```{r}
#| label: check-model-diagostics
#| fig-width: 10
#| fig-heigh: 10
check_model(wb_glm)
```

Read `?check_model` to understand what the different diagnostics produced here show and how to interpret them. Note that these plots are using simulated quantile residuals from DHARMa, which are essential for binary data.

:::